---
subtitle: "TMA4268 Statistical Learning V2019"
title: "Compulsory exercise 1: Group 13"
author: "Erik Andrè Klepp Vik and Vemund Tjessem"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document
  #pdf_document
---
  
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,strip.white=TRUE,prompt=FALSE,
                      cache=TRUE, size="scriptsize",fig.width=4, fig.height=3)
```

```{r,eval=TRUE,echo=FALSE}
# install.packages("knitr") #probably already installed
# install.packages("rmarkdown") #probably already installed
# install.packages("ggplot2") #plotting with ggplot
# install.packages("ggfortify")  
# install.packages("MASS")  
# install.packages("dplyr")  
library(knitr)
library(rmarkdown)
library(ggplot2)
library(ggfortify)
library(MASS)
library(dplyr)
```


# Problem 1

For this problem you will need to include some LaTex code. Please install latex on your computer and then consult Compulsor1.Rmd for hints how to write formulas in LaTex.

## a) 

The expression for the test mean squared error is
\begin{equation}
\text{MSE}_{\text{test}}=\frac{1}{n_0}\sum_{j=1}^{n_0} (y_{0j}-\hat{f}(x_{0j}))^2
\end{equation}

The expected test mean squared error (MSE) at $x_0$ is defined as: 

\begin{equation}
\text{E}[y_0 - \hat{f}(x_0)]^2
\end{equation}

## b)

Using the fact that $y_0=f(x_0)+\varepsilon$ gives

\begin{equation}
\text{E}[y_0 - \hat{f}(x_0)]^2 =  \underbrace{\text{Var}(\varepsilon)}_{\text{Irreducible error}} + \underbrace{\text{Var}(\hat{f}(x_0))}_{\text{Variance of prediction}} + \underbrace{\left( f(x_0) - \text{E}[\hat{f}(x_0)] \right)^2}_{\text{Squared bias}}
\end{equation}

## c)

## d)

## e)

## f)

## g)

# Problem 2

Here is a code chunk:

```{r, eval=TRUE}
id <- "1nLen1ckdnX4P9n8ShZeU7zbXpLc7qiwt" # google file ID
d.worm <- read.csv(sprintf("https://docs.google.com/uc?id=%s&export=download", id))
head(d.worm)
```




## a)


## b) 
Below you have to complete the code and then replace `eval=FALSE` by `eval=TRUE` in the chunk options:
```{r, eval=FALSE, echo=TRUE}
ggplot(d.worm,aes(x= ... ,y=  ... ,colour= ...)) + geom_point() + theme_bw()
```

Note that the default figure width and height have been set globally as `fig.width=4, fig.height=3`, but if you would like to change that (e.g., due to space constraints), you can include a different width and height directly into the chunk options, again using `fig.width=..., fig.height=...`.

## c)

## d)

## e)

## f)

## g)


# Problem 3

## a)

## b)

## c)

## d)

## e)

## f)

## g)

## h)


# Problem 4

## a)

## b)

## c)

## d)





